A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-26 paper.tex
Move SPR section content elsewhere

@@

\section*{Relationship between ARGs and trees}

We have seen that the local tree for a given position on
the genome can be extracted from an ARG by starting at the
leaves and following the appropriate paths through the graph.
However, this tells us little about the relationship between
adjacent trees, and what the properties of overall ensemble of
local trees are.
A recombination between two adjacent trees causes them to
differ by a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the ``subtree transfer'' operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

The effects of recombination as we move from left-to-right along
the genome in an unsimplified gARG are easy to interpret.
Consider the highlighted tree in Fig~\ref{fig-simplification}A.
In it the parent of \textsf{e} is \textsf{j}, whereas in the next tree
its parent is \textsf{k}. Since both parents are present as
unary nodes in the tree, we simply change one edge in
order to effect the transition.
% Need revising to fit new example
A more complex transition occurs when we have a regraft operation
that occurs above the current root (i.e., the current oldest
non-unary node). For example, the transition between the
second and third trees corresponds to node D switching
from the path through I to H, and we then must add branches
corresponding to passing through L and meeting at M.

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is important to note that while extracting local
trees from an ARG is straightforward, the converse
process of reconstructing precisely the same ARG from
these trees is more problematic. If the internal
nodes are consistently labelled and unary nodes are
included in the local trees (e.g., Fig~\ref{fig-simplification})
then it is easy to exactly reconstruct the ARG.
Otherwise, the problem is much more difficult and it
may not be possible to uniquely reconstruct the original ARG.
Without labels for the internal nodes, we are forced to
to reason about which subtrees remain the same and which
differ due to the SPR separating adjacent trees in order to
assign node identity. Without the internal unary nodes marking
recombinations, we must infer their existence. In
general, there will not be a unique solution to this problem,
and so there can be many possible ARGs compatible with
a given list of leaf-labelled binary trees.
This is the approach taken by the Espalier inference
method~\citep{rasmussen2022espalier}, which finds a
set of SPRs that reconcile independently inferred
trees from different regions of the genome.

>>>>

% Needs another pass
An important consequence of simplifying local trees to remove
unary nodes is that we may lose the ability to identify
the ``subtree prune-and-regraft'' (SPR) operations
by which they differ~\citep{hein1990reconstructing,song2003on,song2006properties}.
% The effects of recombination as we move from left-to-right along
% the genome in the complete simulated gARG
% of Fig~\ref{fig-simplification}A are easy to interpret,
% but as more redundancy is removed from the graph
% the
As explained above, we can have multiple recombination and
common ancestor events stacking up, and therefore
many possible SPR operations. Deducing the SPRs between
trees is NP-hard in
general~\citep{hein1996complexity,allen2001subtree,bordewich2005computational},
but effective heuristics have been proposed~\citep{rasmussen2022espalier}.

% The equivalence between an ARG and a set of local trees
% separated by SPRs is often mentioned [e.g. x y z], but
% it is important to note that while extracting local
% trees from an ARG is straightforward, the converse
% process of reconstructing precisely the same ARG from
% these trees is more problematic. If the internal
% nodes are consistently labelled and unary nodes are
% included in the local trees (e.g., Fig~\ref{fig-simplification})
% then it is easy to exactly reconstruct the ARG.
% Otherwise, the problem is much more difficult and it
% may not be possible to uniquely reconstruct the original ARG.
% Without labels for the internal nodes, we are forced to
% to reason about which subtrees remain the same and which
% differ due to the SPR separating adjacent trees in order to
% assign node identity. Without the internal unary nodes marking
% recombinations, we must infer their existence. In
% general, there will not be a unique solution to this problem,
% and so there can be many possible ARGs compatible with
% a given list of leaf-labelled binary trees.
% This is the approach taken by the Espalier inference
% method~\citep{rasmussen2022espalier}, which finds a
% set of SPRs that reconcile independently inferred
% trees from different regions of the genome.

@@

% The equivalence between an ARG and a set of local trees
% separated by SPRs is often mentioned [e.g. x y z], but
% it is important to note that while extracting local
% trees from an ARG is straightforward, the converse
% process of reconstructing precisely the same ARG from
% these trees is more problematic. If the internal
% nodes are consistently labelled and unary nodes are
% included in the local trees (e.g., Fig~\ref{fig-simplification})
% then it is easy to exactly reconstruct the ARG.
% Otherwise, the problem is much more difficult and it
% may not be possible to uniquely reconstruct the original ARG.
% Without labels for the internal nodes, we are forced to
% to reason about which subtrees remain the same and which
% differ due to the SPR separating adjacent trees in order to
% assign node identity. Without the internal unary nodes marking
% recombinations, we must infer their existence. In
% general, there will not be a unique solution to this problem,
% and so there can be many possible ARGs compatible with
% a given list of leaf-labelled binary trees.
% This is the approach taken by the Espalier inference
% method~\citep{rasmussen2022espalier}, which finds a
% set of SPRs that reconcile independently inferred
% trees from different regions of the genome.
